Distributional Trees for Circular Data

### 2019-01-31 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
further revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2417 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 42f98be05571c49204ae08d5ef330278e73d2db0

% Brief abstract of the paper:
\abstract{\textit{TO DO}}

@@

\section{Motivation}
Circular response variables occur in a variety of applications and subject areas. 
For example, gun crime data on a $24$-hour scale is analysed in the social-economics, 
animal orientation or gene-structure analysis are often subject of examination in biology, 
and wind direction is one of the most important weather variables in meteorology.
Representing such a two-dimensional direction by an angle in a unity circle requires 
the specification of a starting point (angle 0) and either a clockwise or anti-clockwise 
rotation. However, these settings can vary but should not influence statistical 
inference from a fitted model.
For that reason circular data demand for purpose-built statistical methods. 
A commonly chosen distribution for modeling circular data is the von Mises distribution.
Estimating its two distribution parameters by generalized additive models using a set of 
regressor variables is one possible approach that can be successful in various settings.
As an alternative we propose the application of distributional trees 
which are capable of detecting also non-additive effects and provide easily interpretable 
models allowing for a wide range of inference procedures.
Moreover, covariates and their possible interactions are selected automatically
such that no expert knowledge is required to specify the model in advance.

@@

%\subsection{Distributional Trees}
%Distributional trees~(Schlosser et. al, 2018) embed distributional modeling into the framework 
%of regression trees.
Distributional trees~(Schlosser et. al, 2018) embed distributional modeling such as
provided by generalized additive models for location, scale, and shape~(GAMLSS, 
Rigby and Stasinopoulos, 2005) into the framework of regression trees~(Breiman et. al, 1984).
In particular one of the unbiased recursive partitioning algorithms MOB~(Zeileis et. al, 2008) 
and CTree~(Hothorn et. al, 2006) is applied and a distributional model is fitted in each node 
by estimating the parameters of a specified distribution family via maximum likelihood.
For each split the split variable is selected based on $p$-values from permutation or M-fluctuation 
tests evaluating associations between model scores and each possible split variable. After choosing 
the split variable showing the highest association to the goodness of fit of the model the corresponding split point is selected in a separate step by optimizing an objective function.

In the here discussed application the MOB algorithm is applied and the prespecified 
distribution family is the von Mises distribution which is also known as the circular 
normal distribution. Its density function is given by
\begin{equation}
  f_{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
\end{equation}
with a location parameter $\mu$, a dispersion parameter $\kappa$, and $I_0(\kappa)$ being the 
modified Bessel function of the first kind and order $0$.
For $\kappa=0$ a uniform distribution is obtained. Otherwise, a large value of $\kappa$ indicates
a high concentration of the distribution around $\mu$. Hence, $\frac{1}{\kappa}$ is analogous
to the variance $\sigma^2$ of the normal distribution. 
(See e.g., Jammalamadaka and Sengupta (2001) or Mardia and Jupp (2009) for a more detailed overview.)
The distribution parameter vector $\theta = (\mu,\kappa)$ is estimated 
by maximizing $l(y, \mu, \kappa) = log(f_{vM}(y;\mu, \kappa))$ and model scores are calculated 
by 
$s(y,\hat{\mu}, \hat{\kappa}) = \left(\frac{\partial l(y, \hat{\mu}, \hat{\kappa}}{\partial \mu},
\frac{\partial l(y, \hat{\mu}, \hat{\kappa}}{\partial \kappa}\right)$. 
%$s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$. 
In that way
a value indicating the goodness of fit is provided for each parameter and observation
such that changes can be detected not only in the location parameter $\mu$ but also in the 
dispersion parameter $\kappa$.

@@

\begin{figure}[!ht]\centering
\includegraphics[width = 1.\textwidth]{circtree_plot_small.pdf}
\caption{\label{fig:circtree} Fitted distribution tree for a circular response employing the von Mises distribution.}\label{fig:tree}
\end{figure}

@@

%Innsbruck lies with the west-east oriented Inn Valley with an approx. 2300\,m
%high mountain ridge on the northern side and with the south-north oriented Wipp
%Valley between the approx. 2300\,m high mountains on the southern side.
%Therefore, skill full covariates are pressure differences along the Inn Valley
%as indices for up- or down-valley winds. For the typical dry gap wind coming
%from the south, indices are, e.g. low relative humidity or temperature
%differences between Innsbruck and stations up the Wipp Valley.

@@

While the application of the von Mises distribution allows for modeling a circular response variable,
additional adaptions have to be made when including circular regressor variables in a distributional tree
model. In particular, if the selected split variable is a circular variable this demands for finding
not only one split point but a ``split interval''. This consideration is of interest for many applications,
for example when modeling wind directions it might be desirable to include day time or earlier 
observations of wind directions as regressor variables.

@@

\item[Breiman, L., Friedman, J. H., Olshen, R. A., and Stone, C. J.] (1984).
     Classification and Regression Trees.
     {\it Wadsworth}.

@@

\item[Mardia, K. V., and Jupp, P. E.] (2009).
     Directional Statistics.
     {\it John Wiley \& Sons}, {\bf 494}.

\item[Rigby, R. A., and Stasinopoulos, D. M.] (2005).
     Generalized Additive Models for Location Scale and Shape (with Discussion).
     {\it Journal of the Royal Statistical Society C}, {\bf 54}, 3,
     507\,--\,554.

\item[Schlosser, L., Hothorn, T., and Zeileis, A.] (2018).
     Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.
     {\it arXiv.org E-Print Archive}, arXiv 1804.02921, \url{http://arxiv.org/abs/1804.02921}

>>>>

% Brief abstract of the paper:
\abstract{Circular response variables occur in a variety of applications, e.g., when
geographical directions are considered or measurements are taken on a $24$-hour scale. 
A commonly chosen distribution for modeling circular data is the von Mises distribution.
As an alternative to estimating its distribution parameters by generalized additive models 
we propose the application of distributional trees which are capable of detecting also 
non-additive effects, select covariates and their possible interactions automatically and 
provide easily interpretable models allowing for a wide range of inference procedures. 
This approach is illustrated in an application to wind directions with a set of meteorological
covariates, all measured at the airport of Innsbruck, Austria.}

%which are capable of detecting also non-additive effects and provide easily interpretable 
%models allowing for a wide range of inference procedures.
%Moreover, covariates and their possible interactions are selected automatically
%such that no expert knowledge is required to specify the model in advance.

@@

%\section{Motivation}
%Circular response variables occur in a variety of applications and subject areas. 
%For example, gun crime data on a $24$-hour scale is analysed in the social-economics, 
%animal orientation or gene-structure analysis are often subject of examination in biology, 
%and wind direction is one of the most important weather variables in meteorology.
%Representing such a two-dimensional direction by an angle in a unity circle requires 
%the specification of a starting point (angle 0) and either a clockwise or anti-clockwise 
%rotation. However, these settings can vary but should not influence statistical 
%inference from a fitted model.
%For that reason circular data demand for purpose-built statistical methods. 
%A commonly chosen distribution for modeling circular data is the von Mises distribution.
%Estimating its two distribution parameters by generalized additive models using a set of 
%regressor variables is one possible approach that can be successful in various settings.
%As an alternative we propose the application of distributional trees 
%which are capable of detecting also non-additive effects and provide easily interpretable 
%models allowing for a wide range of inference procedures.
%Moreover, covariates and their possible interactions are selected automatically
%such that no expert knowledge is required to specify the model in advance.

@@

%\subsection{Distributional Trees}
Distributional trees~(Schlosser et. al, 2018) embed distributional modeling into the framework 
of regression trees.
%Distributional trees~(Schlosser et. al, 2018) embed distributional modeling such as
%provided by generalized additive models for location, scale, and shape~(GAMLSS, 
%Rigby and Stasinopoulos, 2005) into the framework of regression trees~(Breiman et. al, 1984).
In particular one of the unbiased recursive partitioning algorithms MOB~(Zeileis et. al, 2008) 
and CTree~(Hothorn et. al, 2006) is applied and a distributional model is fitted in each node 
by estimating the parameters of a specified distribution family via maximum likelihood.
For each split the split variable is selected based on $p$-values from permutation or M-fluctuation 
tests evaluating associations between model scores and each possible split variable. After choosing 
the split variable showing the highest association to the goodness of fit of the model the corresponding split point is selected in a separate step by optimizing an objective function.

In the here discussed application the MOB algorithm is applied and the prespecified 
distribution family is the von Mises distribution which is also known as the circular 
normal distribution. Its density function is given by
\begin{equation}
  f_{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
\end{equation}
with location parameter $\mu$, dispersion parameter $\kappa$, and $I_0(\kappa)$ being the 
modified Bessel function of the first kind and order $0$.
%For $\kappa=0$ a uniform distribution is obtained. Otherwise, a large value of $\kappa$ indicates
%a high concentration of the distribution around $\mu$. Hence, $\frac{1}{\kappa}$ is analogous
%to the variance $\sigma^2$ of the normal distribution. 
(See e.g., Jammalamadaka and Sengupta (2001) for a more detailed overview.)

In each node the distribution parameters $\mu$ and $\kappa$ are 
%vector $\theta = (\mu,\kappa)$ is 
estimated by maximizing $l(y, \mu, \kappa) = log(f_{vM}(y;\mu, \kappa))$ and model scores are calculated by 
$s(y,\hat{\mu}, \hat{\kappa}) = \left(\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \mu},
\frac{\partial l(y, \hat{\mu}, \hat{\kappa})}{\partial \kappa}\right)$. 
%$s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$. 
In that way a value indicating the goodness of fit is provided for each parameter and observation
such that changes can be detected not only in the location parameter $\mu$ but also in the 
dispersion parameter $\kappa$.

@@

%\begin{figure}
%\includegraphics[height = 0.5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
%\caption{\label{fig:circtree} Fitted distribution tree for a circular response employing the von Mises %distribution.}\label{fig:tree}
%\end{figure}

@@

Innsbruck lies with the west-east oriented Inn Valley with an approx. 2300\,m
high mountain ridge on the northern side and with the south-north oriented Wipp
Valley between the approx. 2300\,m high mountains on the southern side.
Therefore, skill full covariates are pressure differences along the Inn Valley
as indices for up- or down-valley winds. For the typical dry gap wind coming
from the south, indices are, e.g. low relative humidity or temperature
differences between Innsbruck and stations up the Wipp Valley.

@@

This application shows that including covariates such as day time or earlier observations 
of wind directions as (unordered) factors provides valuable information. However, in order 
to properly account not only for the circularity of the response but also of regressors, 
additional adaptions of distributional trees are required. 
In particular, if the selected split variable is a circular variable this demands for 
finding a ``split interval'' instead of a split point.
%While the application of the von Mises distribution allows for modeling a circular response variable,
%additional adaptions have to be made when including circular regressor variables in a distributional tree
%model. In particular, if the selected split variable is a circular variable this demands for finding
%not only one split point but a ``split interval''. This consideration is of interest for many applications,
%for example when modeling wind directions it might be desirable to include day time or earlier 
%observations of wind directions as regressor variables.

@@

%\item[Breiman, L., Friedman, J. H., Olshen, R. A., and Stone, C. J.] (1984).
%     Classification and Regression Trees.
%     {\it Wadsworth}.

@@

%\item[Mardia, K. V., and Jupp, P. E.] (2009).
%     Directional Statistics.
%     {\it John Wiley \& Sons}, {\bf 494}.

%\item[Rigby, R. A., and Stasinopoulos, D. M.] (2005).
%     Generalized Additive Models for Location Scale and Shape (with Discussion).
%     {\it Journal of the Royal Statistical Society C}, {\bf 54}, 3,
%     507\,--\,554.

\item[Schlosser, L., Hothorn, T., and Zeileis, A.] (2018).
     Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.
     {\it arXiv.org E-Print Archive}, https://arxiv.org/abs/1804.02921v2.
